---
source: Rmd
title: "Estimating the mean, variance and standard deviation"
objectives:
  - "Estimate the mean, variance and standard deviation of a variable through simulation."
keypoints:
  - The mean is the average value expected in the population. The mean is also known as the expectation.
  - The variance is the average of the squared differences between values in the distribution and the mean of the distribution.
  - The standard deviation is the square root of the variance. It measures the spread of observations around the mean, in units of the original data. 
questions:
  - How are the mean, variance and standard deviation calculated and interpreted?
teaching: 10
execises: 10
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
#source("../bin/obtain_data.R")
load("../bin/data.RData")
knitr_fig_path("01-")
library(ggplot2)
library(dplyr)
```

Often when doing statistics, we have a variable of interest, for which we want to
estimate particular properties. A variable will follow a *distribution*, which shows
what values a variable can take and how likely these are to occur.
In this episode we will learn to estimate and 
interpret the mean, variance and standard deviation of a variable's distribution. 
These values allow us to estimate the average of a variable in the population and the 
variation around that average. 

Before discussing the definitions of these values, let's look at an example. The
heights of people of the female sex in the US approximately follow a distribution 
with a mean of 162 cm and a standard deviation of 6.89 cm. 
The distribution is shown below. Values on the x-axis with a greater density on the
y-axis have a higher chance of occuring. So a height of 162 cm is most likely to occur, while a height of
140 cm is very unlikely to occur. 

```{r height density plot, echo = FALSE}
range <- tibble(limits = c(162 - 6.89 * 4, 162 + 6.89 * 4))

ggplot(range, aes(x=limits)) +
 stat_function(fun = dnorm,
               args = list(mean = 162,
                           sd = 6.89)) +
 xlab("Height (cm)") +
 ylab("Density")
```

You may recognise the shape of the above distribution. It is an example of a
*normal* distribution. 

Let's say we sampled 1000 observations of female height. 
We can refer to these observations as $y_i$, with $i = 1, \dots, 1000$. 
The distribution of female heights has a few properties of interest, which we can estimate using our sample.

## The mean
The *mean* of the distribution is the average of all the values that make up that distribution. In the case of female heights in the US, the mean is 162 cm. Therefore, the average height of a US female is 162 cm. We also expect the mean of our sample to equal 162 cm.

The mean is also known as the *expectation* of a variable, $E()$. The expectation of female heights is expressed as $E(y)$:

$$E(y) = \frac{1}{n} \left( \sum_{i=1}^n x_{i} \right)$$

where $n = 1000$.

## The variance
The *variance* is the average *squared* difference between values in the distribution 
and the mean of the distribution. This is a mouthful, so it is useful to look
at the equation of variance. The variance is expressed as $V()$:  

$$V(y) = E( (y - E(y))^2 )$$

Breaking this down, we see that the variance is calculated using:
* $y - E(y)$, i.e. the difference between an observed height and the mean height.  
* $(y - E(y))^2$, i.e. the squared difference between an observed height and the mean height.  
* $E( (y - E(y))^2 )$, i.e. the expectation of this squared difference.

Why are we interested in this value? We are mainly interested in the variance
because it allows us to calculate the standard deviation, which can be interpreted
on the original scale. Let's look at this below.

## The standard deviation
The *standard deviation* of a distribution is the *square root* of the variance.
The standard deviation is expressed as $\sigma_y$:

$$\sigma_y = \sqrt{V(y)}$$

The standard deviation is interpreted as a measure of the difference between
values in the distribution and the mean of the distribution. A higher standard deviation
indicates that the spread around the mean is greater. There is no "good" or "bad"
standard deviation - its purpose is to give us an idea of the spread of observations
in the population. 

## Estimating the mean, variance and standard deviation through simulation
Another route to understanding the mean, variance and standard deviation is
through *simulation*. This is the process of *sampling* data from a distribution
using R. This is akin to collecting observations in the real world, where
obervations come from an underlying distribution. 

For example, we can sample 1000 observations from the distribution
of female heights in the US using `rnorm()`. We store these values in a tibble,
with the column named `heights`.

```{r set seed heights, echo = FALSE}
set.seed(112)
```


```{r sample from female heights}
sample <- tibble(heights = rnorm(1000, mean = 162, sd = 6.89))
```

Before calculating the mean, variance and standard deviation, let's create
a histogram of our sample. Although not a perfect representation, our histogram
looks quite similar to the distribution shown at the start of this episode. Your
plot will differ slightly from the histogram shown below, as `rnorm()` obtained
a random sample. If you create a new `sample` and run the `ggplot()` code again, 
the histogram will differ slightly again. Every `sample` contains 1000 new observations
from the same distribution, akin to running an experiment where we collect 1000
observations. 

```{r histogram sample female heights, message = FALSE}
ggplot(sample, aes(x = heights)) +
  geom_histogram()
```

We can calculate the mean of our sample using `mean()`. This value lies 
close to the original mean of 162 cm. Here we see that the mean of a distribution
approximately equals the mean of our sample.

```{r mean female heights}
meanHeight <- mean(sample$heights)
meanHeight
```

We can also calculate the variance as the mean of the squared differences
between our sampled observations and the mean of our sample:

```{r variance female heights}
variance <- mean( (sample$heights - meanHeight)^2 )
variance
```

Finally, we calculate the standard deviation as the square root of the variance.
The square root is obtained using `sqrt()`. Recall that we calculate the standard
deviation to have a measure of spread in our distribution, in the same units
as our original data (in this case, mmHg).

```{r square root female heights}
sqrt(variance)
```


> ## Exercise
> The normal distribution of systolic blood pressure has a mean of 112 mmHg 
> and a standard deviation of 10 mmHg. The distribution looks as follows:
> 
> ```{r set seed BP, echo = FALSE}
> set.seed(112)
> ```
> 
> ```{r blood pressure density plot, echo = FALSE}
> range <- tibble(limits = c(112 - 10 * 4, 112 + 10 * 4))
> 
> ggplot(range, aes(x=limits)) +
>  stat_function(fun = dnorm,
>                args = list(mean = 112,
>                            sd = 10)) +
>  xlab("Systolic blood pressure (mmHg)") +
>  ylab("Density")
> ```
> A) Sample one thousand observations from this distribution. Then, create a histogram of your sample.  
> B) Calculate the average systolic blood pressure in your sample.
> Does this value correspond to the mean of the distribution?    
> C) Calculate the variance of your sample.  
> D) Calculate the standard deviation of your sample. How is the standard deviation interpreted here?
> 
> > ## Solution
> > Throughout this solution, your results will differ slightly from the ones
> > shown below. This is a consequence of `rnorm()` drawing random samples.
> > If you are completing this episode in a workshop setting, ask your 
> > neighbour to compare results! If you are working through this episode
> > independently, try running your code again to see how the results differ.
> > 
> > A) We obtain 1000 observations from the systolic blood pressure distribution using `rnorm()`. 
> > We store these values in an object named `sample`. We then create a histogram
> > using `geom_histogram()`.
> > 
> > ```{r blood pressure sample and histogram, message = FALSE}
> > sample <- tibble(bloodPressure = rnorm(1000, mean = 112, sd = 10))
> > 
> > ggplot(sample, aes(x = bloodPressure)) +
> >   geom_histogram()
> > ```
> > 
> > B) We can then calculate the average systolic blood pressure 
> > in our sample using `mean()`. The mean of our sample lies closely to the mean of the original distribution. 
> > 
> > ```{r estimate mean blood pressure}
> > meanBP <- mean(sample$bloodPressure)
> > meanBP
> > ```
> > 
> > C) The variance equals the average of the squared differences between the mean of the distribution and the
> > observed values. This can be calculated as follows:
> > 
> > ```{r estimate variance blood pressure}
> > variance <- mean( (sample$bloodPressure - meanBP)^2 )
> > variance
> > ```
> > 
> > D) The standard deviation equals the square root of the variance. 
> > It is interpreted as a measure of the spread of sampled systolic blood pressure 
> > values and the mean systolic blood pressure. The square root can be calculated using `sqrt()`. 
> > 
> > ```{r estimate standard deviation blood pressure}
> > sqrt(variance)
> > ```
> {: .solution}
{: .challenge}

