---
source: Rmd
title: "Estimating the variation around the mean: standard errors and confidence intervals"
objectives:
  - "Explore the difference between a population parameter and sample estimates by simulating data from a normal distribution."
  - "Define the 95% confidence interval by simulating data from a normal distribution."
keypoints:
  - "Sample means are expected to differ from the population mean. We quantify the spread of estimated means around the population mean using the standard error."
  - "95% of 95% confidence intervals are expected to capture the population mean. In practice, this means we are fairly confident that a 95% confidence interval will contain the population mean, but we do not know for sure."
questions:
  - "What are the definitions of the standard error and the 95% confidence interval?"
  - "How is the 95% confidence interval interpreted in practice?"
teaching: 20
exercises: 40
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
#source("../bin/obtain_data.R")
load("../bin/data.RData")
knitr_fig_path("02-")
library(ggplot2)
library(dplyr)
```

In the previous episode, we obtained an estimate of the mean by simulating from
a normal distribution. This simulation was akin to obtaining 
one sample in the real world. We learned that as is the case in the real
world, every sample will differ from the next sample slightly.

In this episode, we will learn how to estimate the variation in the mean estimate.
This will tell us how far we can expect the population mean to lie from our sample mean. 
We will also visualise the variation in the
estimate of the mean, when we simulate many times from the same distribution.
This will allow us to understand the meaning of the 95% confidence interval.

## The standard error 
In the previous episode, we learnt that the standard deviation is a measure of the 
expected difference between a value in the distribution and the mean of
the distribution. In this episode we will work with a related quantity, called
the *standard error*. The standard error quantifies the spread of sample means
around the population mean. 

For example, let's say we collected 1000 observations of female heights in
the US, to estimate the mean height. The standard error of this estimate in this case is 
`r round(6.89/sqrt(1000), 2)` cm. 

The standard error is calculated using the standard deviation and the sample size:

$$\text{se}(y) = \sigma_y/\sqrt{n}.$$

Therefore, for our female heights sample, the standard error was calculated
as $\text{se}(y) = 6.89 / \sqrt{1000} \approx 0.22$ cm.

Notice that with greater $n$, the standard error will be smaller. This makes
intuitive sense: with greater sample sizes come more precise estimates of the mean.

One of the reasons we calculate the standard error is that it allows us to
calculate the 95% confidence interval, as we will do in the next section. 

## The 95% confidence interval
We use the standard error to calculate the 95% confidence interval. 
This interval equals, for large samples:

$$[E(y) - 1.96 \times \text{se}(y); E(y) + 1.96 \times \text{se}(y)],$$

i.e. the mean +/- 1.96 times the standard error.

The definition of the 95% confidence interval is a bit convoluted: 

*95% of 95% confidence intervals are expected to contain the true population mean*.

Imagine we were to obtain 100 samples of female heights in the US, with each
sample containing 1000 observations. For each sample, 
we would calculate a mean and a 95% confidence interval. By the definition of
the interval, we would expect approximately 95 of the intervals to contain the true mean of 162 cm.
In other words, we would expect approximately 5 of our samples to give confidence intervals that
do not contain the true mean of 162 cm.

What does this mean in practice? When we calculate a 95% confidence interval, 
we are fairly confident that it will contain the true population mean, but
we cannot know for sure. 

Let's take a look at how we can calculate the 95% confidence interval in R.
First, we sample 1000 observations of female heights and calculate the mean, 
as in the previous episode. Then, we calculate the standard error as the
standard deviation divided by the square root of the sample size.
Finally, the 95% confidence interval is given by the mean +/- 1.96
times the standard error. Note that your CI will differ slightly from the
one shown below, as `rnorm()` obtains random samples. 

```{r set seed heights 1, echo = FALSE}
set.seed(689)
```

```{r heights standard error and CI}
sample <- tibble(heights = rnorm(1000, mean = 162, sd = 6.89))

meanHeight <- mean(sample$heights)

seHeight <- 6.89/sqrt(1000)

CI <- c(meanHeight - 1.96 * seHeight, meanHeight + 1.96 * seHeight)
CI
```

The confidence interval shown above has a lower bound of 161.7898 
and an upper bound of 162.6439. In this instance, we have obtained
a 95% confidence interval that captures the population mean of 162.
The confidence interval that you obtain may not capture the population mean
of 162. It would then be part of the 5% of 95% confidence intervals
that do not capture the population mean. 

> ## Exercise
> A) Given the distribution of systolic blood pressure, with a mean of 112 mmHg
> and a standard deviation of 10 mmHg, calculate the standard error of a mean
> estimate coming from a sample of 2000 observations. How do we interpret this
> value?  
> B) Using a random sample of 2000 observations from this distribution, estimate
> the mean systolic blood pressure and provide a 95% confidence interval for this
> estimate. How do we interpret this interval?
> 
> > ## Solution
> > Throughout this solution, your results will differ slightly from the ones
> > shown below. This is a consequence of `rnorm()` drawing random samples.
> > If you are completing this episode in a workshop setting, ask your 
> > neighbour to compare results! If you are working through this episode
> > independently, try running your code again to see how the results differ.
> > 
> > A) The standard error equals the standard deviation divided by the square root
> > of the sample size. We expect the difference between the estimated mean
> > and the population mean to equal 0.22 mmHg, on average. 
> > ```{r blood pressure standard error}
> > seBP <- 10/sqrt(2000)
> > seBP
> > ```
> > 
> > B) We simulate using `rnorm()`, calculate the mean using `mean()` and calculate
> > the confidence interval as the mean +- 1.96 times the standard error. Since
> > 95% of 95% confidence intervals are expected to contain the true population mean, we are
> > fairly confident that the interval that we have calculated contains the true
> > population mean (but we are not sure, as 5% of 95% confidence intervals 
> > are not expected to contain the true population mean).
> > ```{r blood pressure confidence interval}
> > sample <- tibble(bloodPressure = rnorm(1000, mean = 112, sd = 10))
> > meanBP <- mean(sample$bloodPressure)
> > 
> > CI <- c(meanBP - 1.96 * seBP, meanBP + 1.96 * seBP)
> > CI
> > ```
> > The confidence interval has lower bound 111.1055 and upper bound
> > 111.9821. The interval does not capture the population mean of 
> > 112 and therefore belongs to the 5% of 95% CIs that do not capture
> > the population mean. 
> {: .solution}
{: .challenge}

## Simulating 95% confidence intervals
In the previous section we learned that 95% of 95% confidence intervals
are expected to contain the true population mean. Let's see this in 
action, by simulating 100 data sets of 1000 female heights.

First, we create a tibble for our means and confidence intervals.
We name this tibble `means` and create empty columns for the sample IDs,
mean heights, the lower bound of the confidence intervals and the upper
bound of the confidence intervals. 

```{r set seed heights 2, echo = FALSE}
set.seed(689)
```

```{r tibble for heights data}
means <- tibble(sampleID = numeric(),
               meanHeight = numeric(),
               lower_CI = numeric(),
               upper_CI = numeric())
```

Then, we calculate the standard error using the standard deviation and the sample size.

```{r standard error heights data}
seHeight <- 6.89/sqrt(1000)
```

We simulate the Heights data using a `for` loop. This gives us 100 iterations,
in which samples are drawn as before using `rnorm()`. For each sample, 
we add a row to our `means` tibble using `add_row()`. We include an ID for the
sample, the mean Height and the bounds of the 95% confidence interval. 

```{r simulate heights data}
for(i in 1:100){
 sample <- tibble(heights = rnorm(1000, mean = 162, sd = 6.89))
 
 means <- means %>%
   add_row(sampleID = i,
           meanHeight = mean(sample$heights),
           lower_CI = mean(sample$heights) - 1.96 * seHeight,
           upper_CI = mean(sample$heights) + 1.96 * seHeight)
}
```

Finally, we add a column to our `means` tibble that indicates whether a 
confidence interval contains the true population mean of 162 cm.

```{r add capture column to heights data}
means <- means %>%
 mutate(capture = ifelse((lower_CI < 162 & upper_CI > 162),
                           "Yes", "No"))
```

Now we are ready to plot our means and confidence intervals. The following
code plots the mean estimates along the x-axis using `geom_point()`, with
sample IDs along the y-axis. Confidence intervals are added to these points
using `geom_errorbar()`. The population mean is displayed as a dashed line
using `geom_vline()`. Finally, confidence intervals are coloured by whether
they capture the true population mean. 

```{r plot heights confidence intervals}
ggplot(means, aes(x = meanHeight, y = sampleID, colour = capture)) +
 geom_point() +
 geom_errorbar(aes(xmin = lower_CI, xmax = upper_CI, colour = capture)) +
 geom_vline(xintercept = 162, linetype = "dashed") +
 scale_color_manual("Population mean captured",
                    values = c("red", "black"))
```

In this instance, 2 out of 100 95% confidence intervals did not capture
the population mean. If you run the above code multiple times, you will
discover that there is variation in the number of confidence intervals
that capture the population mean. On average, 5 out of 100 95% confidence
intervals will not capture the population mean. 

> ## Exercise
> A) Given the distribution of systolic blood pressure, with a mean of 112 mmHg
> and a standard deviation of 10 mmHg, simulate 100 data sets of 2000 observations
> each. For each sample, store the sample ID, mean systolic blood pressure and the 
> 95% confidence interval in a tibble.  
> B) Create an extra column in the tibble, indicating for each confidence
> interval whether the original mean of 112 mmHg was captured.  
> C) Plot the mean values and their 95% confidence intervals. Add a line to show the population mean. How
> many of your 95% confidence intervals do not contain the true population 
> mean of 112 mmHg? 
> 
> > ## Solution
> > Throughout this solution, your results will differ slightly from the ones
> > shown below. This is a consequence of `rnorm()` drawing random samples.
> > If you are completing this episode in a workshop setting, ask your 
> > neighbour to compare results! If you are working through this episode
> > independently, try running your code again to see how the results differ.
> > 
> > A) 
> > ```{r set seed blood pressure, echo = FALSE}
> > set.seed(112)
> > ```
> > 
> > ```{r simulate blood pressure data}
> > means <- tibble(sampleID = numeric(),
> >                 meanBP = numeric(),
> >                 lower_CI = numeric(),
> >                 upper_CI = numeric())
> > 
> > seBP <- 10/sqrt(2000)
> > 
> > for(i in 1:100){
> >   
> >   sample <- tibble(bloodPressure = rnorm(2000, mean = 112,
> >                                          sd = 10))
> >   
> >   means <- means %>%
> >     add_row(sampleID = i,
> >             meanBP = mean(sample$bloodPressure),
> >             lower_CI = mean(sample$bloodPressure) - 1.96 * seBP,
> >             upper_CI = mean(sample$bloodPressure) + 1.96 * seBP)
> > }
> > ```
> > 
> > B)
> > ```{r add capture column to blood pressure data}
> > means <- means %>%
> >   mutate(capture = ifelse((lower_CI < 112 & upper_CI > 112),
> >                             "Yes", "No"))
> > ```
> > 
> > C) In this instance, three 95% confidence intervals did not contain
> > the true population mean. Since `rnorm()` draws random samples,
> > you may have more or less 95% confidence intervals that capture the 
> > true population mean. On average, we 5 out of 100 95% confidence
> > intervals will fail to capture the population mean. 
> > 
> > ```{r plot blood pressure confidence intervals}
> > ggplot(means, aes(x = meanBP, y = sampleID, colour = capture)) +
> >   geom_point() +
> >   geom_errorbar(aes(xmin = lower_CI, xmax = upper_CI, colour = capture)) +
> >   geom_vline(xintercept = 112, linetype = "dashed") +
> >   scale_color_manual("Population mean captured",
> >                      values = c("red", "black"))
> > ```
> > 
> {: .solution}
{: .challenge}
