---
source: Rmd
title: "Predicting means using linear associations"
objectives:
  - "Predict the mean of one variable through its association with a binary variable."
  - "Predict the mean of one variable through its association with a continuous variable."
keypoints:
  - "We can calculate means and confidence intervals for a continous outcome variable, grouped by a binary or continuous explanatory variable. On a small scale, this is an example of a model."
questions:
  - "How can the mean of a continous outcome variable be predicted with a binary explanatory variable?"
  - "How can the mean of a continous outcome variable be predicted with a continous explanatory variable?"
teaching: 10
execises: 10
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
#source("../bin/obtain_data.R")
load("../bin/data.RData")
knitr_fig_path("04-")
library(ggplot2)
library(dplyr)
library(tidyr)
```

In this episode we will be predicting the **mean** of one variable through its
linear association with another variable. In a way, this will be our first example
of a *model*: under the assumption that the mean of one variable is associated
with another variable, we will have the capacity to predict that mean. 

This episode will bring together the concepts of *mean*, *confidence interval* and
*association*, covered in the previous episodes. Prediction through linear regression
will then be formalised in the [next lesson](https://carpentries-incubator.github.io/simple-linear-regression-public-health).

We will be referring to the variable for which we are making predictions as the 
*outcome* variable. The variable used to make predictions will be referred to
as the *explanatory* variable. 

## Mean prediction using a binary explanatory variable
We will start with using a binary explanatory variable, i.e. a variable which
can take on one of two values. For example, we may try to predict mean
systolic blood pressure (`BPSysAve`) by physical activity (`PhysActive`). 
Physical activity is a binary variable in the NHANES data, as it is coded
as "Yes" or "No". 

Let's first explore the association between systolic blood pressure and 
physical activity. This can be done using a violin plot, which shows
the density distribution of systolic blood pressure by physical activity.

```{r BPSysAve PhysActive violin plot}
dat %>%
  drop_na(BPSysAve, PhysActive) %>%
  ggplot(aes(x = PhysActive, y = BPSysAve)) +
  geom_violin() +
  ylab("Systolic blood pressure") +
  xlab("Physically active")
```

It is hard to tell from the above plot whether the mean systolic blood pressure
differs across physical activity. Let's calculate the means and 
confidence intervals. In the code below, we group observations by physical
activity using `group_by()`. We then calculate the mean, standard error and
confidence interval bounds inside `summarise()`.

```{r BPSysAve PhysActive means}
means <- dat %>%
  drop_na(c(BPSysAve, PhysActive)) %>%
  group_by(PhysActive) %>%
  summarise(
    mean = mean(BPSysAve),
    n = n(),
    se = sd(BPSysAve) / sqrt(n()),
    lower_CI = mean - 1.96 * se,
    upper_CI = mean + 1.96 * se)

means
```
It appears that the mean estimate for systolic blood pressure is approximately 
5 units lower for physically active participants than for non-physically
active participants. 

Now we can overlay the mean estimates and their confidence intervals onto
the violin plot. First the means are overlayed using `geom_point()`. Then 
the confidence intervals are overlayed using `geom_errorbar()`.

```{r BPSysAve PhysActive violin with mean}
dat %>%
  drop_na(c(BPSysAve, PhysActive)) %>%
  ggplot(aes(x = PhysActive, y = BPSysAve)) +
  geom_violin() +
  geom_point(data = means, aes(x = PhysActive, y = mean)) +
  geom_errorbar(data = means, aes(x = PhysActive, y = mean, ymin = lower_CI, ymax = upper_CI),
                width = 0.2) +
  ylab("Systolic blood pressure") +
  xlab("Physically active")
```

On a small scale, we have created a model of *mean* systolic blood pressure
by physical activity. Hereby we have obtained mean estimates and confidence 
intervals for systolic blood pressure as a function of physical activity. 

> ## Exercise
> In this exercise you will explore the association between total HDL cholesterol (`TotChol`) and
> smoking (`SmokeNow`). Ensure that you drop NAs from `SmokeNow` and `TotChol`, by including
> `drop_na(SmokeNow, TotChol)` in your piped commands.
> 
> A) Create a violin plot of total cholesterol by smoking status.  
> B) Calculate the mean total cholesterol by smoking status, along with the 95% confidence 
> interval for this mean estimate.  
> C) Overlay these mean estimates and their confidence intervals onto the violin plot.
> 
> > ## Solution
> > A) 
> > ```{r TotChol smokenow violin plot}
> > dat %>%
> >   drop_na(SmokeNow, TotChol) %>%
> >   ggplot(aes(x = SmokeNow, y = TotChol)) +
> >   geom_violin()
> > ```
> > 
> > B) 
> > ```{r FEV1 smokenow means and CIs}
> > means <- dat %>%
> >   drop_na(c(SmokeNow, TotChol)) %>%
> >   group_by(SmokeNow) %>%
> >   summarise(
> >     mean = mean(TotChol),
> >     n = n(),
> >     se = sd(TotChol) / sqrt(n()),
> >     lower_CI = mean - 1.96 * se,
> >     upper_CI = mean + 1.96 * se)
> > ```
> > 
> > C) 
> > ```{r FEV1 smokenow violin with mean}
> > dat %>%
> >   drop_na(c(SmokeNow, TotChol)) %>%
> >   ggplot(aes(x = SmokeNow, y = TotChol)) +
> >   geom_violin() +
> >   geom_point(data = means, aes(x = SmokeNow, y = mean)) +
> >   geom_errorbar(data = means, aes(x = SmokeNow, y = mean, ymin = lower_CI, ymax = upper_CI),
> >                 width = 0.2) +
> >   ylab("Total HDL cholesterol") +
> >   xlab("Currently smoking")
> > ```
> {: .solution}
{: .challenge}

## Mean prediction using a continuous explanatory variable
Now we will expand the approach learnt above to prediction using a 
continuous explanatory variable. We will try to predict
Weight from Height in adult participants. 

Let's first explore the association between Height and Weight. To make grouping
observations easier, we will round Height to the nearest integer
before plotting and performing downstream calculations. We will round
using `mutate()`, as shown in the plotting command below.

```{r explore association rounded Height and Weight, warning = FALSE}
dat %>%
  drop_na(Weight, Height) %>%
  filter(Age > 17) %>%
  mutate(Height = round(Height)) %>%
  ggplot(aes(x = Height, y = Weight)) +
  geom_point()
```

Now we can calculate the mean Weight by Height:

```{r mean Weight by Height}
means <- dat %>%
  drop_na(Weight, Height) %>%
  filter(Age > 17) %>%
  mutate(Height = round(Height)) %>%
  group_by(Height) %>%
  summarise(
    mean = mean(Weight),
    n = n(),
    se = sd(Weight) / sqrt(n()),
    lower_CI = mean - 1.96 * se,
    upper_CI = mean + 1.96 * se
  )
```

Finally, we can overlay these means and confidence intervals onto the 
scatterplot. 

```{r explore association rounded Height and Weight with means, warning = FALSE}
dat %>%
  drop_na(Weight, Height) %>%
  filter(Age > 17) %>%
  mutate(Height = round(Height)) %>%
  ggplot(aes(x = Height, y = Weight)) +
  geom_point() +
  geom_point(data = means, aes(x = Height, y = mean), colour = "red", size = 2) +
  geom_errorbar(data = means, aes(x = Height, y = mean, ymin = lower_CI, ymax = upper_CI),
                colour = "red")
```

We see that as the positive correlation coefficient suggested, mean
Weight indeed increases with Height. The outer confidence intervals are much
wider than the central confidence intervals, because many less observations
were used to estimate the outer means. There are also a few red points
without a confidence interval, because those means were estimated with single
observations only. 

> ## Exercise
> In this exercise you will explore the association between total FEV1 (`FEV1`) and
> age (`Age`). Ensure that you drop NAs from `FEV1` by including
> `drop_na(FEV1)` in your piped commands. Also make sure to filter for adult
> participants by including `filter(Age > 17)`.
> 
> A) Create a scatterplot of FEV1 as a function of age.   
> B) Calculate the mean FEV1 by age, along with the 95% confidence 
> interval for these mean estimates.  
> C) Overlay these mean estimates and their confidence intervals onto the scatterplot.
> 
> > ## Solution
> > A)
> > ```{r FEV1 age scatterplot, warning = FALSE}
> > dat %>%
> >   drop_na(FEV1) %>%
> >   filter(Age > 17) %>%
> >   ggplot(aes(x = Age, y = FEV1)) +
> >   geom_point() 
> > ```
> > 
> > B) 
> > ```{r mean FEV1 by age}
> > means <- dat %>%
> >   drop_na(FEV1) %>%
> >   filter(Age > 17) %>%
> >   group_by(Age) %>%
> >   summarise(
> >     mean = mean(FEV1),
> >     n = n(),
> >     se = sd(FEV1) / sqrt(n()),
> >     lower_CI = mean - 1.96 * se,
> >     upper_CI = mean + 1.96 * se
> >   )
> > ```
> > 
> > C) 
> > ```{r FEV1 age scatterplot with means, warning = FALSE}
> > dat %>%
> >   drop_na(FEV1) %>%
> >   filter(Age > 17) %>%
> >   ggplot(aes(x = Age, y = FEV1)) +
> >   geom_point() +
> >   geom_point(data = means, aes(x = Age, y = mean), colour = "red", size = 2) +
> >   geom_errorbar(data = means, aes(x = Age, y = mean, ymin = lower_CI, ymax = upper_CI),
> >                 colour = "red")
> > ```
> {: .solution}
{: .challenge}