---
source: Rmd
title: "Predicting means using linear associations"
objectives:
  - "Predict the mean of one variable through its association with a continuous variable."
keypoints:
  - "We can predict the means, and calculate confidence intervals, of a continuous outcome variable grouped by a continuous explanatory variable. On a small scale, this is an example of a model."
questions:
  - "How can the mean of a continous outcome variable be predicted with a continous explanatory variable?"
teaching: 10
execises: 10
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
#source("../bin/obtain_data.R")
load("../bin/data.RData")
knitr_fig_path("04-")
library(ggplot2)
library(dplyr)
library(tidyr)
```

In this episode we will predict the **mean** of one variable using its
linear association with another variable. In a way, this will be our first example
of a *model*: under the assumption that the mean of one variable is associated
with another variable, we will be able to predict that mean. 

This episode will bring together the concepts of *mean*, *confidence interval* and
*association*, covered in the previous episodes. Prediction through linear regression
will then be formalised in the [next lesson](https://carpentries-incubator.github.io/simple-linear-regression-public-health).

We will refer to the variable for which we are making predictions as the 
*outcome* variable. The variable used to make predictions will be referred to
as the *explanatory* variable. 

## Mean prediction using a continuous explanatory variable
We will predict the mean of one continuous outcome variable using a 
continuous explanatory variable. Specifically, we will try to predict
mean Weight using Height of adult participants. 

Let's first explore the association between Height and Weight. To make predicting easier,
we will group observations together: we will round Height to the nearest integer
before plotting and performing downstream calculations. We will round
using `mutate()`, as shown in the plotting command below.

```{r explore association rounded Height and Weight, warning = FALSE}
dat %>%
  drop_na(Weight, Height) %>%
  filter(Age > 17) %>%
  mutate(Height = round(Height)) %>%
  ggplot(aes(x = Height, y = Weight)) +
  geom_point()
```

Now we can calculate the mean Weight for each Height. In the code below, 
we first remove rows with missing 
values using `drop_na()` from the `tidyr` package. 
We then filter for adult participants using `filter()` and round Heights using
`mutate()`. Next, we group observations by rounded height
using `group_by()`. We finally calculate the mean, standard error and
confidence interval bounds using `summarise()`.

```{r mean Weight by Height}
means <- dat %>%
  drop_na(Weight, Height) %>%
  filter(Age > 17) %>%
  mutate(Height = round(Height)) %>%
  group_by(Height) %>%
  summarise(
    mean = mean(Weight),
    n = n(),
    se = sd(Weight) / sqrt(n()),
    lower_CI = mean - 1.96 * se,
    upper_CI = mean + 1.96 * se
  )
```

Finally, we can overlay these means and confidence intervals onto the 
scatterplot. First the means are overlayed using `geom_point()`. 
Then the confidence intervals are overlayed using `geom_errorbar()`.

```{r explore association rounded Height and Weight with means, warning = FALSE}
dat %>%
  drop_na(Weight, Height) %>%
  filter(Age > 17) %>%
  mutate(Height = round(Height)) %>%
  ggplot(aes(x = Height, y = Weight)) +
  geom_point() +
  geom_point(data = means, aes(x = Height, y = mean), colour = "red", size = 2) +
  geom_errorbar(data = means, aes(x = Height, y = mean, ymin = lower_CI, ymax = upper_CI),
                colour = "red")
```

We see that as the positive correlation coefficient suggested, mean
Weight indeed increases with Height. The outer confidence intervals are much
wider than the central confidence intervals, because many fewer observations
were used to estimate the outer means. There are also a few red points
without a confidence interval, because those means were estimated using only a single
observation. 

> ## Exercise
> In this exercise you will explore the association between total FEV1 (`FEV1`) and
> age (`Age`). Ensure that you drop NAs from `FEV1` by including
> `drop_na(FEV1)` in your piped commands. Also, make sure to filter for adult
> participants by including `filter(Age > 17)`.
> 
> A) Create a scatterplot of FEV1 as a function of age.   
> B) Calculate the mean FEV1 by age, along with the 95% confidence 
> interval for each of these mean estimates.  
> C) Overlay these mean estimates and their confidence intervals onto the scatterplot.
> 
> > ## Solution
> > A)
> > ```{r FEV1 age scatterplot, warning = FALSE}
> > dat %>%
> >   drop_na(FEV1) %>%
> >   filter(Age > 17) %>%
> >   ggplot(aes(x = Age, y = FEV1)) +
> >   geom_point() 
> > ```
> > 
> > B) 
> > ```{r mean FEV1 by age}
> > means <- dat %>%
> >   drop_na(FEV1) %>%
> >   filter(Age > 17) %>%
> >   group_by(Age) %>%
> >   summarise(
> >     mean = mean(FEV1),
> >     n = n(),
> >     se = sd(FEV1) / sqrt(n()),
> >     lower_CI = mean - 1.96 * se,
> >     upper_CI = mean + 1.96 * se
> >   )
> > ```
> > 
> > C) 
> > ```{r FEV1 age scatterplot with means, warning = FALSE}
> > dat %>%
> >   drop_na(FEV1) %>%
> >   filter(Age > 17) %>%
> >   ggplot(aes(x = Age, y = FEV1)) +
> >   geom_point() +
> >   geom_point(data = means, aes(x = Age, y = mean), colour = "red", size = 2) +
> >   geom_errorbar(data = means, aes(x = Age, y = mean, ymin = lower_CI, ymax = upper_CI),
> >                 colour = "red")
> > ```
> {: .solution}
{: .challenge}
